---
title: "Leaf Color Change Date in Harvard Forest"
author: "Kevin Buck, Kyle Elliott, Sydney Higgins"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#plotting
library(ggplot2)
# data manipulation
library(tidyverse)
# making residual plots
library(ggResidpanel)
# estimating marginal means
library(emmeans)
# needed for creating proper ANOVA table
library(car)
#Anova stat package
library(rstatix)
library(ggpubr)
```

####Introduction####
Researcher John O'Keefe and his team starting in 1991 collected various statistics for a number of species of trees in the Prospect Hill Tract Harvard Forest in Petersham, MA. He has continued to record the statistics and compile them in an online data archive in order to provide researchers with data to investigate if certain trends are correlated with certain species. In this data archive, one particular data set has statistics regarding the mean day of the year in which each tree of various given species has 50% of their leaves change color. We will be investigating how this day (listed in Julian days within the data set) is correlated with species type and investigate the trends for each species throughout the last 30 years. 

Our initial hypothesis is that this would be the defining factor because the trees are different species. Still, because of climate change, we expect warmer temperatures to last longer. This impacts the leaf color-changing process, which we predict will have a net result of later Julian days for the 50% quota to be reached by each species. Leaf color change affects nutrient cycling, so understanding the relationship between climate change and leaves changing color in different species is crucial for predicting if the long-term effects of climate change or different species are the dominant factor when leaves change. If climate change is the defining factor, we would see a time-dependent differentiating change from year to year. In contrast, if species were the defining factor, we would see an asymmetric time-dependent differentiating change that isn't dependent on the year-to-year as it would be on each species.
###Hypotheses###
Null: There is no species or year effect on leaf color change date. It will be random with noise
HA1: The leaf color change date would increase each year with no species dependence.
HA2: The leaf color change date would be different for each species, but not change from year to year.
HA3:The leaf color change date would be later as years progress, with each species' leaf color change date being different as well.
###Data Acquisition###
O'Keefe and his team's data was collected through weekly observations of percent leaf coloration and percent leaf fall started in September each year. Researchers observed when 50% of the leaves of trees of different species had changed color and started to fall, and recorded those dates. They recorded this data for 33 unique types of species from 1991 to the present day, with the most recent data coming from 2022.

####Analysis####
Upon piping our desired data set, we filtered down to the 14 species that had recorded mean Julian dates for 50% of the leaves changing colors for each of the tree species. To assess the changes in the Julian date of leaf color change in each species per year, we first created a scatter plot of the Julian date vs. year and had each species of tree which were represented by the different colors. We added trend lines for each species to better visualize the overall trend for each species.

To further investigate species and their overall trend in 50% leaf color changes as a mean Julian day, we created another graph in which we grouped the species within the same genus to see if trees of a more similar genetic background have similar leaf color changing effects. This graph will be able to help us determine if there is a specific impact seen by the species type and if the trends are correlated within each genus of tree or if they are specific to each species.

```{r}
pheno <- read.csv(file="hf003-08-fall-mean-spp.csv") #read in the data
pheno_trimmed <- pheno %>% #pipe in data
  #filter in species present for the whole dataset
  filter(species=="ACRU"|species=="ACSA"|species=="ACPE"|species=="AMSP"|
           species=="BEAL"|species=="BELE"|species=="BEPA"|species=="FAGR"|
           species=="FRAM"|species=="NYSY"|species=="PRSE"|species=="QUAL"|
           species=="QURU"|species=="QUVE")

pheno_trimmed %>% 
  ggplot(aes(x = year, y = lc_doy)) + #set year as x axis and julian day of change as the y
  geom_point(aes(color=species, fill=species), size=1) + 
  ggtitle("Leaf Color Change Date Over Time")+ #add title
  #adding in trend lines by species
  geom_smooth(aes(group=species,color=species), method = "lm", formula = y ~ x, se = F)+
  xlab("Year") + #label x axis
  ylab("Date 50% of Leaves Changed Color ") + #label y axis
  theme_classic() #changing background
```
ANCOVA
```{r}
options(contrasts = c("contr.sum","contr.ply")) #options fo later post hoc comparison
model_1 <- lm(lc_doy ~ year * species,data=pheno_trimmed) #build linear model
summary(model_1) #view linear model
resid_panel(model_1,plot="all") #check on residuals
Anova(model_1, type = 3) #run an ANCOVA
pwc <- pheno_trimmed %>% tukey_hsd(lc_doy ~ species) #pair wise comparisons
pwc #look at the PWC of species


```

###Results###
What do graphs show
Anova

###Conclusion###
Based on the analysis, we can conclude leaf turn date is more climate dependent than species-dependent, but species-dependent is still significant. We also see the interaction between species and year is significant, which means different species aren't responding in the same way to climate changes as years go on. This is not what we hypothesized, but makes intuitive sense, as different trees respond differently to changing seasons. Further testing will help clarify what effect species differences affect the leaf drop date, and help us understand which species are more or less affected by climate change from year to year. One of the further tests we can do is compare different genera and different species within genera to see if the leaf change date is more dependent on species or genus.


###Contribution Statement###
We extend our appreciation to everyone who made this project possible, specifically Mr. O'Keefe and his team at Harvard that collected the data that we were able to use as well as Dr. McLachlan and his team of TAs (specifically Gretchen and Eva) that helped to advise and educate us on biostatistics. This project would not have been possible without your work as well, so we thank you for making this opportunity for our group!

###Citations###
O'Keefe J. 2023. Phenology of Woody Species at Harvard Forest since 1990. Harvard Forest Data Archive: HF003 (v.35). Environmental Data Initiative: https://doi.org/10.6073/pasta/eb0dd36c6ec62a918340b6bda38be832.
