---
title: "Leaf Color Change Date in Harvard Forest"
author: "Kevin Buck, Kyle Elliott, Sydney Higgins"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#plotting
library(ggplot2)
# data manipulation
library(tidyverse)
# making residual plots
library(ggResidpanel)
# estimating marginal means
library(emmeans)
# needed for creating proper ANOVA table
library(car)
#Anova stat package
library(rstatix)
library(ggpubr)
```

####Introduction####
Researcher John O'Keefe and his team starting in 1991 collected various statistics for a number of species of trees in the Prospect Hill Tract Harvard Forest in Petersham, MA. He has continued to record the statistics and compile them in an online data archive in order to provide researchers with data to investigate if certain trends are correlated with certain species. 
In this data archive, one particular data set has statistics regarding the mean day of the year in which each tree of various given species has 50% of their leaves change color. We will be investigating how this day (listed in Meridian days within the data set) is correlated with species type and investigate the trends for each species throughout the last 30 years. 

###Data Acquisition###
O'Keefe and his team's data was collected through weekly observations of percent leaf coloration and percent leaf fall started in September each year. Researchers observed when 50% of the leaves of trees of different species had changed color and started to fall, and recorded those dates. They recorded this data for 33 unique types of species from 1991 to the present day, with the most recent data coming from 2022.

####Analysis####
Upon piping our desired data set, we filtered down to the 14 species that had recorded mean meridian dates for 50% of the leaves changing colors for each of the tree species. To assess the changes in the meridian date of leaf color change in each species per year, we first created a scatter plot of the meridian date vs. year and had each species of tree which were represented by the different colors. 

how did we choose to analyze the data with our given question
Data visualization

```{r}
pheno <- read.csv(file="hf003-08-fall-mean-spp.csv") #read in the data
pheno_trimmed <- pheno %>% #pipe in data
  #filter in species present for the whole dataset
  filter(species=="ACRU"|species=="ACSA"|species=="ACPE"|species=="AMSP"|
           species=="BEAL"|species=="BELE"|species=="BEPA"|species=="FAGR"|
           species=="FRAM"|species=="NYSY"|species=="PRSE"|species=="QUAL"|
           species=="QURU"|species=="QUVE")

pheno_trimmed %>% 
  ggplot(aes(x = year, y = lc_doy)) + #set year as x axis and julian day of change as the y
  geom_point(aes(color=species, fill=species), size=1) + 
  ggtitle("Leaf Color Change Date Over Time")+ #add title
  #adding in trend lines by species
  geom_smooth(aes(group=species,color=species), method = "lm", formula = y ~ x, se = F)+
  xlab("Year") + #label x axis
  ylab("Date 50% of Leaves Changed Color ") + #label y axis
  theme_classic() #changing background
```
ANCOVA
```{r}
options(contrasts = c("contr.sum","contr.ply")) #options fo later post hoc comparison
model_1 <- lm(lc_doy ~ year * species,data=pheno_trimmed) #build linear model
summary(model_1) #view linear model
resid_panel(model_1,plot="all") #check on residuals
Anova(model_1, type = 3) #run an ANCOVA
pwc <- pheno_trimmed %>% tukey_hsd(lc_doy ~ species) #pair wise comparisons
pwc #look at the PWC of species

#pwc %>% 
  #ggplot(aes(x = species, y = lc_doy)) + #set year as x axis and julian day of change as the y
  #geom_bar(aes(color=species, fill=species), size=2) + 
  #ggtitle("Leaf Color Change Date Over Time")+ #add title
  #xlab("Species") + #label x axis
  #ylab("Date 50% of Leaves Changed Color ") + #label y axis
  #theme_classic() #changing background
```

###Results###
What do graphs show
Anova

###Conclusion###
Conclusion from anova

###Contribution Statement###
We extend our appreciation to everyone who made this project possible, specifically Mr. O'Keefe and his team at Harvard that collected the data that we were able to use as well as Dr. McLachlan and his team of TAs that helped to advise and educate us on biostatistics. This project would not have been possible without your work as well, so we thank you for making this opportunity for our group!

###Citations###
O'Keefe J. 2023. Phenology of Woody Species at Harvard Forest since 1990. Harvard Forest Data Archive: HF003 (v.35). Environmental Data Initiative: https://doi.org/10.6073/pasta/eb0dd36c6ec62a918340b6bda38be832.
